---
title: "Using R to Explore The Impact of COVID-19 on Venice's Airbnb Rates"
author: "Matt Purtill"
date: "20/01/2021"
output:
   md_document:
      preserve_yaml: TRUE
always_allow_html: true

---

```{r setup, include=FALSE}
# Initial setup and opening up libraries

knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(leaflet)
library(htmlwidgets)


```

Prior to 2020 Venice was used as one of the most harmful examples of overtourism in the world. Like all tourist hotspots, the impact of short-term rentals on the local housing market has been under scrutiny, with the threat of regulation over Airbnb-style home rentals.

The coronavirus pandemic changed everything, as international tourism during the 2020. This article will examine the impact of the pandemic on Venice's Airbnb ecosystem between August 2019 and August 2020. We will consider how the distribution and cost of properties changed in the year, and which locations have bee n the most impacted.

## Our Data Source

Data for this analysis is obtained from [Inside Airbnb](http://insideairbnb.com/). This site offers accessible downloads on Airbnb property data from major cities around the world. This gives us in-depth data on each Airbnb property in Venice at the time of the data scrape. Although it does not impact the data, it is worth noting that the site takes a critical stance towards Airbnb's impact on cities, and provides its own analysis on featured cities.

This analysis will be constructed using R, using mostly tidyverse ggplot visualisations and the leaflet mapping package. and an R Markdown copy of the code is available here.


```{r load data and data cleaning, include = FALSE}

listings_2019 <- read.csv("2019_listings.csv", stringsAsFactors = FALSE)

listings_2020 <- read.csv("2020_listings.csv", stringsAsFactors = FALSE)

# We can combine the datasets into one data table

listings <- listings_2019 %>%
  mutate(year = 2019) %>%
  bind_rows(listings_2020) %>%
  mutate(year = ifelse(is.na(year), 2020, 2019))

listings %>%
  filter(year==2019) %>%
  nrow()

listings <- listings %>%
  mutate(price = as.numeric(gsub('[$,]','',price))) %>%
  mutate_if(is.character, factor)

```

## Venice's Pre-Coronavirus Airbnb Ecosystem

```{r summary statistics, include = FALSE}

# How many listings were there in August 2019

listings %>%
  filter(year == 2019) %>%
  summarise(
    properties = n(),
    adr = mean(price),
    lowest_adr = min(price),
    highest_adr = max(price)
  )

```

In August 2019 there are 8,907 properties listed for Venice with an ADR of \$139. However, we can see a massive range in average rates - from a low of \$9 a night to a suspicious value of \$12,345. We can build a histogram to display the spread of ADR across properties.

```{r ADR histogram, echo = FALSE}

# Building a histogram of all properties from 2019

listings %>%
  filter(year == 2019) %>%
  ggplot(aes(x=price)) +
  geom_histogram(bins = 60) +
  theme_bw() +
  labs(y="Count of Properties", x="Average Daily Rate (USD)")

```

Lots of the definition in the low range is lost because of the few very high ADR properties. As most of the action is in the \$1000 or below range, we can do an alternative histogram for properties that are in the price range of us mere-mortals. As we are looking at a price variable, I will also compare the distribution with a logarithmic value to see if it brings to a more symmetical distribution.

```{r ADR histogram elaborated, echo = FALSE}

# First a histogram with 2019 properties and with daily rate less than 1000 dollars a day

listings %>%
  filter(year == 2019, price < 1000) %>%
  ggplot(aes(x=price)) +
  geom_histogram(bins = 60) +
  theme_bw() +
  labs(y="Count of Properties", x="Average Daily Rate (USD)")

# The same histogram as above, but with logged prices to obtain more normal distribution.

listings %>%
  filter(year == 2019, price < 1000) %>%
  ggplot(aes(x=log(price))) +
  geom_histogram(bins = 60) +
  theme_bw() +
  labs(y="Count of Properties", x="Log of Average Daily Rate (USD)")

```

As we might expect, the ADR is distributed witsh a positive skew, which becomes more symmetrical with a logarithmic function. This will be useful to know for modelling later on.

Of course, not all properties are equal. The city of Venice includes the well-known central islands, but also includes some more peripheral islands, as well as a large area of the mainland, which connects to the islands by bridge. We can plot the median daily rate for each neighbourhood of the city to show how much of a premium is placed on the a position in the islands. Unsurprisingly, the island neighbourhoods also show the largest number of properties on short-term rental sites.


```{r choropleth of 2019 prices, echo = FALSE}

# Creating a summary table that will be joined with a GeoJSON file
# Some additional summary statistics are calculated but not used - number of properties and 
# % that are entire rooms or apartments
# NB neighbourhood_cleansed is the field with the lowest-level of usable neighbourhood for mapping. 

nhood_prices_last_year <- listings %>%
  filter(year == 2019) %>%
  group_by(neighbourhood_cleansed) %>%
  summarise(
    price = median(price),
    properties = n(),
    perc_entire_homes = sum(room_type == "Entire home/apt")/n()
            ) %>%
  rename("neighbourhood" = "neighbourhood_cleansed")

# Reading GeoJSON file of Venice neighbourhoods - the original file seems to give the
# incomplete patchwork of the city

neighbourhoods <- rgdal::readOGR("neighbourhoods.geojson")

# Joining our summary stats table to the map object

neighbourhoods@data <- left_join(neighbourhoods@data, nhood_prices_last_year, by="neighbourhood")

# Setting a palette for our choropleth map

pal <- colorNumeric("magma", domain = NULL, reverse = TRUE)

# Building the map with tiles, polygons and formatting labels

leaflet(neighbourhoods) %>%
  addProviderTiles(providers$Esri.WorldGrayCanvas) %>%
  setView(lng = 12.3155, 
          lat = 45.4408, 
          zoom = 11) %>%
  addPolygons(stroke = FALSE,
              smoothFactor = 0.3, 
              fillOpacity = 0.8, 
              fillColor = ~pal(price),
              label = ~ifelse(is.na(price),
                              paste0(neighbourhood, ": No Properties"),
                              paste0(neighbourhood," , Average Rate: $",price))
  ) %>%
  addLegend(pal = pal, values = ~price,
            position = "bottomright", 
            title= "Median ADR",
            labFormat = labelFormat(prefix = "$")) %>%
  saveWidget(here::here('','neighbourhood_map.html')) # save map widget

```

<iframe src="neighbourhood_map.html" height="600px" width="100%" style="border:none;"></iframe>



## Impact of COVID-19 on Summer Bookings

- stays/availability
- impact on prices on like-for-like properties

```{r comparing year-on-year, echo = FALSE}

listings %>%
  group_by(year) %>%
  summarise(
    properties = n(),
    price = median(price),
  ) %>%
  pivot_wider(names_from = year,
              values_from = c(properties, price))

```

There are approximately 400 fewer properties available in Venice, and median daily rates are down by $17 (-15%).

```{r comparing neighbourhoods, echo = FALSE}

# Building a summary dataset grouped by neighbourhood and year for median price and no. of properties

# Then we are pivoting wider and comparing the property number and price change on a percentage basis

nhood_cv_impact <- listings %>%
  group_by(neighbourhood_cleansed, neighbourhood_group_cleansed, year) %>%
  summarise(
    price = median(price),
    properties = n()
            ) %>%
  pivot_wider(names_from = year,
              values_from = c(price, properties)) %>%
  rename("neighbourhood" = "neighbourhood_cleansed") %>%
  mutate(yoy_change_prop = (properties_2020 - properties_2019)/properties_2019,
         yoy_change_rate = (price_2020 - price_2019)/price_2019)

# Plotting each neighbourhood by change in price and change in count of properties

# Colour of points and labels corresponds to the island/mainland location of the neighbourhood

nhood_cv_impact %>%
  filter(properties_2020 > 40) %>%
  ggplot(aes(x=yoy_change_prop, y=yoy_change_rate)) +
  geom_point(aes(color=neighbourhood_group_cleansed)) +
  theme_bw() +
  ggrepel::geom_label_repel(aes(label = 
                                  ifelse(neighbourhood %in%  c("Lido","Favaro", "San Marco",
                                                   "Sant'Elena", "Cannaregio", "Castello", 
                                                   "Bissuola", "Carpenedo"),
                                         as.character(neighbourhood), ""),
                                color=neighbourhood_group_cleansed), size=3) +
  labs(x = "% Change in Number of Properties vs. 2019",
       y = "% Change in Median Daily Rate vs. 2019") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_color_discrete(name = "Location")


```

Looking at the regional difference in rates and properties and the difference between the islands ('Isole') and the mainland ('Terrafirma'). We can see that the largest impact was felt in the islands. The largest reduction in rates came from the most expensive regions of San Marco, Cannaregio and Castello. There is almost a discrete difference in the rate changes for regions in the islands and the regions on the mainland. We may expect that the mainland regions are less impacted by rates because of their budget base prices. Other economic activity, other than leisure tourism, may also have provided a more stable customer base for mainland rentals.

Lido is the one island region with an increase in rates year-on-year. Lido may have fared well year-on-year because of its position as the city's summer beach resort on the edge of the lagoon. The region may have been boosted by increased numbers of domestic breaks during the pandemic. Additionally, the increase in rates may be due to slightly changing dates from when the data was scraped - the 2019 rates are from early August, whereas the 2020 rates were scraped towards late August just ahead of the popular Ferragosto holiday, when these beach properties would be at their highest demand.

## 

```{r direct impact on same listings, echo = FALSE}

# Dataset grouping the price and availability metrics for properties that appear both in the 2019 and 2020 listings

# We omit properties that are not included in both years with the final filter function


direct_imp <- listings %>%
  group_by(id, year, room_type, neighbourhood_group_cleansed, latitude, longitude) %>%
  summarise(
    price = median(price),
    availability = median(availability_90)
            ) %>%
  pivot_wider(names_from = year,
              values_from = c(price, availability)) %>%
  mutate(yoy_change_availability = (availability_2020 - availability_2019)/90,
         yoy_change_rate = (price_2020 - price_2019)/price_2019) %>%
  filter(!is.na(yoy_change_rate))

# Charting this data into a two-pane boxplots. One pane for Island properties and one on the mainland

# Separate boxplots for private rooms and entire homes/apartments

direct_imp %>%
  filter(yoy_change_rate < 1, room_type!="Shared room") %>%
  ggplot(aes(x=room_type, y=yoy_change_rate)) +
  geom_boxplot(aes(fill = room_type)) +
  facet_wrap(~neighbourhood_group_cleansed, nrow = 2) +
  scale_y_continuous(label = scales::percent) +
  coord_flip() +
  theme_bw() +
  labs(x = "Room Type", y = "% Rate Change vs. 2019") +
  guides(fill = FALSE)

```

We can also track individual properties that have been listed over the past year and use a boxplot to compare the difference in rates between the islands and the mainland. We can also take into account the type of property - whether it is a whole property or a private room in a larger building, which we would expect to have an impact on rate dynamics.

In the islands there is a similar average reduction in rate across both types of property. In these properties, the median reduction in rates is 12%. However, on the mainland it is only in private rooms where there has been a median reduction rate year-on-year. Among entire properties on the mainland, median rates have remained the same despite the coronavirus pandemic. It is worth commenting that in all groups there is a wide range of rate changes, partly due to the individual circumstances of each property, but is perhaps an indication that there is no single strategy that Airbnb hosts has taken during the pandemic.


``` {r impact on central island rents, echo = FALSE}

# We need to setting a second palette for this map because of the diverging property of percentage changes

pal2 <- colorNumeric("RdBu", domain = NULL, reverse = FALSE)

# Plotting all properties on the main islands as a circle marker

direct_imp %>%
  filter(yoy_change_rate <= 1) %>%
  leaflet() %>%
  addProviderTiles(providers$Esri.WorldGrayCanvas) %>%
  setView(lng = 12.335, lat = 45.439, zoom = 14) %>%
  addCircleMarkers(
    lng = ~longitude,
    lat = ~latitude,
    color = ~pal2(yoy_change_rate),
    radius = 3,
    fillOpacity = 1,
    stroke = FALSE
  ) %>%
  addLegend(pal = pal2, values = ~round(yoy_change_rate*100,0),
            labFormat = labelFormat(suffix = "%"),
            bins=5, title = "% Change in Rate vs 2019") %>%
  saveWidget(here::here('', 'property_map.html')) # save map widget
  

```
<iframe src="property_map.html" height="600px" width="100%" style="border:none;"></iframe>



Using R's leaflet mappping package, we can dive deeper into the data, plotting each property that appears in 2019 and 2020. The colour of each dot represents the change in daily rate on a year-on-year basis. Focussing on just the historic main islands, there are far more reds than blues, showing how rates have tumbled in the historic centre of the city - although even in these locations there are still a number of properties that have raised their rates.

We might expect that central areas close to tourist attractions are the most impacted by the reduction in the number of visiting tourists. Around these areas there are plenty of properties that cut their rates, but also a fair number of properties that have raised their rates from 2019. This is surprising, and is maybe something to investigate with further research or modelling.

Also of interest is the reduction in rates in the south-east of the main islands. This drop in rates appears to be consistently steep across most properties in this area. As well as the coronavirus pandemic, one additional difference between 2019 and 2020 is that the previous year saw the Venezia Biennale take place at the nearby Giardini della Biennale. With this popular arts event not taking place in 2020, this is another reason why nearby rates may have decreased so much. More broadly, this may be an additional contributing reason why rental rates would be lower in 2020 in the absence of COVID.

## Conclusions and Further Analyses

With this analysis we have begun to dig into some of the drivers for Airbnb rents in Venice. We have been able to confirm some prior assumptions that rates are higher in the historic centre of the city, and we have also seen how different regions of the city have been impacted separately from the coronavirus pandemic. In our final mapping of properties with annual rate changes, we have been able to identify some overall trends on rate based on property location. However, it is clear that there is much to rate changes than location. This analysis has only scraped the surface of what is available in this dataset so there is a large scope for future analysis:

* Using quantitative measures and text descriptions to model rates and rate changes
* Track the impact of amenities and descriptions on property review scores
* Compare the difference in the market for hosts with multiple properties compared to small-scale single unit hosts